Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-03-05 abundance-measurement.Rmd
Minor text improvements

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in artificially constructed ('mock') communities of bacterial cells is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

@@

**Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
More generally, we can identify an higher-order taxon $I$ with a set of species, $\{i \in I\}$.
The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
The efficiency of taxon $I$ equals the weighted average of its constituent species,
\begin{align}
  (\#eq:efficiency-general)
  \text{efficiency}_I(a) 
    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
\end{align}
Thus the efficiency of a higher-level taxon may vary across samples (@mclaren2019cons).
An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \@ref(eq:efficiency-general) is used.
In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.

@@

Despite the diversity of these methods, each can be represented by Equation (\#eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 
\begin{align}
  (\#eq:density-ratio-error)
  \widehat{\text{abun}}_{i}(a) 
  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
\end{align}
The error thus consists of two terms: the ratio of the efficiency between species $i$ and species $r$, and any error in the abundance of the reference species.
The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.

>>>>

The measured ratio between species $i$ and $j$ is given by the ratio of their read counts, 
\begin{align}
  (\#eq:ratio-meas)
  \widehat{\text{ratio}}_{i/j}(a) = \frac{\text{reads}_i(a)}{\text{reads}_j(a)}.
\end{align}
From Equations \@ref(eq:measurement-model) and \@ref(eq:ratio-meas), it follows that 
\begin{align}
  (\#eq:ratio-error)
  \widehat{\text{ratio}}_{i/j}(a)
%  &= \frac{\text{abun}_{i}(a)}{\text{abun}_{j}(a)} 
  &= \text{ratio}_{i/j}(a) \cdot \underbrace{\frac{\text{efficiency}_{i}}{\text{efficiency}_{j}}}_{\substack{\text{constant} \\ \text{fold error}}}.
\end{align}
Taxonomic bias creates a multiplicative measurement error in the ratio that is equal to the ratio in their efficiencies; the error is therefore constant across samples.
For instance, in Figure \@ref(fig:error-proportions), the ratio of Species 3 (with an efficiency of 6) to Species 1 (with an efficiency of 1) is over-estimated by a factor of 6 in both communities despite their varying compositions.
A demonstration in bacterial mock communities is shown in [Figure 3D](https://doi.org/10.7554/eLife.46923.004) of @mclaren2019cons.

@@

**Higher-order taxa:** How does bias affect the proportions and ratios of higher-order taxa, such as a genera, phyla, or even all organisms?
More generally, we can identify a higher-order taxon $I$ with a set of species, $\{i \in I\}$.
The abundance and reads of a taxon $I$ in a sample $a$ equal the sums over its constituent species, $\text{abun}_{I}(a) = \sum_{i \in I}\text{abun}_{i}(a)$ and $\text{reads}_{I}(a) = \sum_{i \in I}\text{reads}_{i}(a)$.
The efficiency of taxon $I$ equals the weighted average of its constituent species,
\begin{align}
  (\#eq:efficiency-general)
  \text{efficiency}_I(a) 
    \equiv \frac{\sum_{i\in I}\text{abun}_i(a)\cdot \text{efficiency}_i}{\text{abun}_I(a)}.
\end{align}
Thus the efficiency of a higher-level taxon may vary across samples (@mclaren2019cons).
An exception is if the constituent species have the same efficiency or always appear in the same abundances relative to each other, in which case the efficiency of the higher-order taxon would stay constant.
As an example, suppose that Species 1 and Species 2 in Figure \@ref(fig:error-proportions) were in the same phylum.
The efficiency of the phylum would then be $\tfrac{1}{2} \cdot 1 + \tfrac{1}{2} \cdot 18 = 9.5$ in Sample 1 and $\tfrac{15}{16} \cdot 1 + \tfrac{1}{16} \cdot 18 \approx 2.1$ in Sample 2.
Equations \@ref(eq:prop-error) and \@ref(eq:ratio-error) continue to describe the measurement error in proportions and ratios involving higher-order taxa, so long as the sample-dependent taxon efficiency defined by Equation \@ref(eq:efficiency-general) is used.
In this way, we can see that both proportions and ratios among higher-order taxa may have inconsistent fold errors.

@@

Despite the diversity of these methods, each can be represented by Equation \@ref(eq:density-ratio-meas), making it straightforward to understand the impact of taxonomic bias in the MGS measurement.
It follows from Equation \@ref(eq:ratio-error) that the measured abundance of species $i$ is 
\begin{align}
  (\#eq:density-ratio-error)
  \widehat{\text{abun}}_{i}(a) 
  = \text{abun}_{i}(a) \cdot \frac{\text{efficiency}_{i}}{\text{efficiency}_{r}}
  \cdot \underbrace{\frac{\widehat{\text{abun}_r}(a)}{\text{abun}_r(a)}}_{\substack{\text{fold error in} \\ \text{ref. species}}}.
  % \cdot \begin{array}{c} \text{fold error in} \\ \widehat{\text{abun}_r}(a) 
\end{align}
The error thus consists of two terms: the ratio of the efficiency between species $i$ and species $r$, and any error in the abundance of the reference species.
The ratio of species efficiencies is constant by assumption; thus the total error is constant if the error in the reference abundance is.
Note, if a higher-order taxon is used as a reference---for example, using targeted measurements of a genus or family---then the efficiency of the reference taxon may vary across samples, leading to non-proportional errors.

### 2022-03-05 conclusion.Rmd
Minor text improvements

We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
In addition, careful pairing of total-abundance and MGS measurement methods may remove most of the bias-driven error in absolute abundances from total-abundance normalization.

Important open questions and future research directions remain.
<!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
Key open questions include 1) determining the extent to which bias is consistent across samples at the species level for a given MGS method and sample type; 2) assessing the validity of post-extraction absolute-abundance controls as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how our assumptions about underlying community dynamics, and in particular what information zero counts provide, affect bias sensitivity.
In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, better statistical tools are needed that are capable of including taxonomic bias and uncertainty in the MGS and supplemental non-MGS measurements.
Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.

Our theoretical framework and example analyses provide a foundation for addressing these open questions and developing a practical experimental protocols and statistical tools.
We look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.

>>>>

We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
In addition, careful pairing of total-abundance and MGS measurement methods can counter the variation in mean efficiency of the MGS measurement, leading to more consistent fold errors across samples.

Important open questions and future research directions remain.
<!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
Key open questions include 1) determining the extent to which bias is consistent across samples for different taxonomic levels, MGS methods, and sample types; 2) assessing the validity of post-extraction measurements as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how different underlying community dynamics, and in particular the source of zero counts in MGS measurements, affect bias sensitivity.
In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, more sophisticated statistical tools are needed to properly account for both taxonomic bias and random variation in the underlying MGS and supplemental (e.g. qPCR) measurements.
Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.

Our theoretical framework and example analyses provide a foundation for addressing these open questions, as well as developing experimental protocols and statistical tools that implement our proposed solutions.
We look to a future in which microbiome researchers regularly choose an appropriate combination of experimental and data-analytic methods that are capable of answering their fundamental question while also accounting for taxonomic bias and other limitations inherent to MGS measurement.
In doing so, we will collectively gain the confidence needed to codify the findings from MGS-based microbiome studies into true scientific knowledge.
